Validation challenges in large-scale tree crown segmentations from remote sensing imagery using Deep Learning: a case study in Germany

### 2025-06-11 Validation_challenges_tree_crown_segmentation/Draft1_PerspectivePaper.qmd
final

---
title: "Validation challenges in large-scale tree crown segmentations from remote sensing imagery using Deep Learning: a case study in Germany"
title-running: Validation challenges in tree crown segmentations
bibliography: bibliography.bib
format:
  quarto-ecml-pdf:
    keep-tex: true
  quarto-ecml-html: default
author:
  - name: Taimur Khan
    orcid: '0000-0001-7833-5474'
    corresponding: true 
    affiliations:
      - name: Helmholtz Centre for Environmental Research (UFZ)
        address: Theodor-Lieser-Str. 4
        city: Halle (Saale)
        postal-code: 04109 
        country: Germany
        url: taimur.khan@ufz.de
        id: south
    url: https://www.ufz.de/index.php?en=49404
  - name: Jasmin Krebs
    orcid: '0009-0003-9330-1231'
    affiliations:
      - name: Leipzig University
        address: Augustusplatz 10 
        city: Leipzig
        country: Germany
        postal-code: 04109 
        url: jk21byxu@studserv.uni-leipzig.de
  - name: Sharad Kumar Gupta
    orcid: '0000-1111-2222-3333'
    affiliations:
      - ref: south
      - name: Secondary European Affiliation
        address: Tiergartenstr. 17
        city: Heidelberg
        country: Germany
        postal-code: 69121 
        url: sharad.gupta@ufz.de
  - name: Jonathan Renkel
    orcid: '0000-1111-2222-3333'
    affiliations:
      - name: Martin Luther University Halle-Wittenberg
        address: Von-Seckendorff-Platz 2
        city: Halle (Saale)
        country: Germany
        postal-code: 06120
        url: jonathan.renkel@geo.uni-halle.de
  - name: Caroline Arnold
    orcid: '0000-0002-9458-1517'
    affiliations:
      - name: Helmholtz-Zentrum Hereon
        address: Max-Planck-Straße 1
        city: Geesthacht
        country: Germany
        postal-code: 21502 
        url: caroline.arnold@hereon.de
  - name: Nils Nölke 
    orcid: '0000-0003-4925-2287'
    affiliations:
      - name: University of Göttingen, Forest Inventory and Remote Sensing
        address: Büsgenweg 5
        city: Göttingen
        country: Germany
        postal-code: 37077 
        url: nils.noelke@forst.uni-goettingen.de

@@

To evaluate the performance and limitations of large-scale crown segmentation and its validation, the DeepTrees model [@khan2025deeptrees] was applied in a one-shot prediction approach using pretrained model weights provided by Freudenberg et al [@freudenberg2022individual] as well as model weights trained in DeepTrees [@khan2025deeptrees], applied to high-resolution multispectral digital orthophoto imagery (DOP20) covering the German federal states of Saxony (SN) and Saxony-Anhalt (ST). SN, covering approximately 18,450 km², and Saxony-Anhalt, spanning around 20,452 km², represent diverse ecological and urban landscapes, ideal for assessing large-scale segmentation model performance (Figure 3 in Appendices). The 4-channel (RGBi) DOP20 imagery, with a spatial resolution of 20 cm per pixel , enables precise tree crown delineation. The DeepTrees model identified approximately 218.7 million individual tree crowns—137.3 million in SN and 81.4 million in ST (Figure 1a). <!--The resulting segmentation dataset has been made publicly available on Zenodo (https://zenodo.org/record/exampledoi).-->

Despite optimized methods, exact matches between predicted and ground-truth crown spreads were low (32%), though expanding the margin of error to ±5 m increased accuracy significantly (89%), highlighting validation challenges. The segmentation model systematically overestimated small crowns (\<6 m) and underestimated large crowns (\>16 m), yet the crown area versus spread relationship remained stable (Figure 1b), indicating consistent segmentations. Validation quality proved crucial; notably, the local inventory (‘Baumkataster’ Halle/Saale) provided broad intervals (5 m), limiting validation precision but still achieving an overall IOU of \~70%.The segmented tree crowns show substantial regional variability in tree distribution, reflecting ecological, topographical, and land-use gradients (Figure 3). To evaluate the performance and limitations of large-scale crown segmentation and its validation, the DeepTrees model[@khan2025deeptrees] was applied in a one-shot prediction approach using pretrained model weights provided by Freudenberg et al [@freudenberg2022individual] as well as our own model weights, applied to high-resolution multispectral digital orthophoto imagery (DOP20) covering the German federal states of Saxony (SN) and Saxony-Anhalt (ST). SN, covering approximately 18,450 km², and ST, spanning around 20,452 km², represent diverse ecological and urban landscapes, ideal for assessing large-scale segmentation model performance (Figure 3 in Appendices). The 4-channel (RGBi) DOP20 imagery, with a spatial resolution of 20 cm per pixel , enables precise tree crown delineation. The DeepTrees model identified approximately 218.7 million individual tree crowns—137.3 million in SN and 81.4 million in ST (Figure 1a).

::: {.column width="100%"}
```{=latex}
\begin{figure}[htbp]
\centering
\includegraphics[width=0.8\textwidth]{figures/figure1.png}
\caption{a. Spatial distribution of individual tree crown segmentation polygons derived from multispectral DOP20 imagery across the German federal states of Sachsen (SN, green) and Sachsen-Anhalt (ST, blue), totaling 218.7 million predicted crowns.
b. Relationship between normalized crown area (x-axis) and estimated crown spread (y-axis) for all segmented trees. A clear positive correlation is observed, although deviations from the expected linear trend highlight segmentation uncertainties, particularly for small and irregularly shaped crowns. This variation underscores the need for improved methods to capture crown geometry and supports the use of allometric checks for quality assessment.}
\end{figure}

@@

Validating large-scale tree crown segmentation models reveals a web of interrelated challenges. Among these challenges is phenology, a moving target: the same forest can look drastically different between leaf-on summer imagery and leaf-off winter scenes. Models trained on a single phenological period often struggle in another, yielding inconsistent segmentation accuracy across the year [@sapkota2024; @liu2024]. For example, a canopy delineation that performs well on lush summer foliage may under-segment sparse autumn crowns or miss bare branches in winter. Such season-driven variability not only degrades model performance but also complicates validation – a one-shot model might appear accurate in one season and fail in the next, raising questions about how and when accuracy should be assessed. Incorporating multi-season data including the key phenological stages: leaf emergence, flowering, fruiting, autumn coloring and leaf fall, during both training and validation is essential, as phenological dynamics have been shown to strongly influence model generalizability [@moussaid2021tree; @cong2022citrus].

@@

::: {.column width="100%"}
```{=latex}
\begin{figure}[htbp]
\centering
\includegraphics[width=0.8\textwidth]{figures/figure2.png}
\caption{Illustration of key validation challenges in large-scale tree crown segmentation across two federal states (Sachsen-Anhalt; in blue \& Sachsen; in green) in Germany. Insets highlight specific issues: (a) seasonal variation (phenology) affecting appearance, (b) tiling artifacts in image preprocessing, (c) misalignment between predicted crowns and known tree stem locations, and (d) discrepancies between predicted crowns (blue) and labeled ground truth (red), emphasizing model performance gaps. Blue (Sachsen-Anhalt) and green (Sachsen) overlays represent modelled individual tree crowns across different Federal States in Germany.}
\end{figure}

@@

Validating large-scale tree segmentation models is challenging due to seasonal variability, diverse terrain and illumination conditions, inconsistent ground truth data, ambiguous evaluation criteria, and scale-dependent performance. These factors significantly affect the reliability and practical value of AI-generated tree maps for forestry, ecology, and urban planning.

@@

\begin{credits}
\subsubsection{\ackname} DeepTrees is part of the [“DeepTrees: Deep-Learning based spatiotemporal tree inventorying and monitoring from public orthoimages”](https://deeptrees.de) project, funded by the Integration Platform "Sustainable Future Land Use" at Helmholtz-Centre for Environmental Research – UFZ within the Programme oriented Funding (PoF) period IV of the Helmholtz Program "Changing Earth – Sustaining our Future", Topic 5 "Landscapes of the Future. This repository is based on the work described in Freudenberg et al. (2022). This work was supported by Helmholtz Association's Initiative and Networking Fund through Helmholtz AI [grant number: ZT-I-PF-5-01]. This work used resources of the Deutsches Klimarechenzentrum (DKRZ) granted by its Scientific Steering Committee (WLA) under project ID AIM.

>>>>

---
title: "Validation challenges in large-scale tree crown segmentations from remote sensing imagery using Deep Learning: a case study in Germany"
title-running: Validation challenges in tree crown segmentations
bibliography: bibliography.bib
no-links: false
format:
  quarto-ecml-pdf:
    keep-tex: true
  quarto-ecml-html: default
author:
  - name: Taimur Khan
    orcid: '0000-0001-7833-5474'
    corresponding: true 
    affiliations:
      - name: Helmholtz Centre for Environmental Research (UFZ)
        address: Theodor-Lieser-Str. 4
        city: Halle (Saale)
        postal-code: 04109 
        country: Germany
        url: taimur.khan@ufz.de
        id: south
  - name: Jasmin Krebs
    orcid: '0009-0003-9330-1231'
    affiliations:
      - name: Leipzig University
        address: Augustusplatz 10 
        city: Leipzig
        country: Germany
        postal-code: 04109 
        url: jk21byxu@studserv.uni-leipzig.de
  - name: Sharad Kumar Gupta
    orcid: '0000-0003-3444-1333'
    affiliations:
      - ref: south
      - name: HZDR - Centre for Advanced System Understanding (CASUS)
        address: Untermarkt 20
        city: Görlitz
        country: Germany
        postal-code: 02826 
        url: sharad.gupta@ufz.de
  - name: Jonathan Renkel
    orcid: '0009-0002-1513-9497'
    affiliations:
      - name: Martin Luther University Halle-Wittenberg
        address: Von-Seckendorff-Platz 4
        city: Halle (Saale)
        country: Germany
        postal-code: 06120
        url: jonathan.renkel@geo.uni-halle.de
  - name: Caroline Arnold
    orcid: '0000-0002-9458-1517'
    affiliations:
      - name: Helmholtz-Zentrum Hereon
        address: Max-Planck-Straße 1
        city: Geesthacht
        country: Germany
        postal-code: 21502 
        url: caroline.arnold@hereon.de
  - name: Nils Nölke 
    orcid: '0000-0003-4925-2287'
    affiliations:
      - name: University of Göttingen, Forest Inventory and Remote Sensing
        address: Büsgenweg 5
        city: Göttingen
        country: Germany
        postal-code: 37077 
        url: nils.noelke@forst.uni-goettingen.de

@@

To evaluate the performance and limitations of large-scale crown segmentation and its validation, the DeepTrees model [@khan2025deeptrees] was applied in a one-shot prediction approach using pretrained model weights provided by Freudenberg et al [@freudenberg2022individual] as well as model weights trained in DeepTrees [@khan2025deeptrees], applied to high-resolution multispectral digital orthophoto imagery (DOP20) covering the German federal states of Saxony (SN) and Saxony-Anhalt (ST) [@lvermgeo2020dop; @geosn_dop20_2022]. SN, covering approximately 18,450 km², and Saxony-Anhalt, spanning around 20,452 km², represent diverse ecological and urban landscapes, ideal for assessing large-scale segmentation model performance (Figure 3 in Appendices). The 4-channel (RGBi) DOP20 imagery, with a spatial resolution of 20 cm per pixel , enables precise tree crown delineation. The DeepTrees model identified approximately 218.7 million individual tree crowns—137.3 million in SN and 81.4 million in ST (Figure 1a). The resulting segmentation dataset has been made available upon request on Zenodo (https://doi.org/10.5281/zenodo.15638573).

Despite optimized methods, exact matches between predicted and ground-truth crown spreads were low (32%), though expanding the margin of error to ±5 m increased accuracy significantly (89%), highlighting validation challenges. The segmentation model systematically overestimated small crowns (\<6 m) and underestimated large crowns (\>16 m), yet the crown area versus spread relationship remained stable (Figure 1b), indicating consistent segmentations. Validation quality proved crucial; notably, the local inventory (‘Baumkataster’ Halle/Saale) provided broad intervals (5 m), limiting validation precision but still achieving an overall IOU of \~70%.The segmented tree crowns show substantial regional variability in tree distribution, reflecting ecological, topographical, and land-use gradients (Figure 3). To evaluate the performance and limitations of large-scale crown segmentation and its validation, the DeepTrees model[@khan2025deeptrees] was applied in a one-shot prediction approach using pretrained model weights provided by Freudenberg et al [@freudenberg2022individual] as well as our own model weights, applied to high-resolution multispectral digital orthophoto imagery (DOP20) covering the German federal states of Saxony (SN) and Saxony-Anhalt (ST). SN, covering approximately 18,450 km², and ST, spanning around 20,452 km², represent diverse ecological and urban landscapes, ideal for assessing large-scale segmentation model performance (Figure 3 in Appendix). The 4-channel (RGBi) DOP20 imagery, with a spatial resolution of 20 cm per pixel, enables precise tree crown delineation. The DeepTrees model identified approximately 218.7 million individual tree crowns—137.3 million in SN and 81.4 million in ST (Figure 1a), and derived allometrical traits of crown area and crown spread (Figure 1b). A clear positive correlation between crown spread and crown area is observed, although deviations from the expected linear trend highlight segmentation uncertainties, particularly for small and irregularly shaped crowns. This variation underscores the need for improved methods to capture crown geometry and supports the use of allometric checks for quality assessment.

::: {.column width="100%"}
```{=latex}
\begin{figure}[htbp]
\centering
\includegraphics[width=0.8\textwidth]{figures/figure1.png}
\caption{(a) Spatial distribution of individual tree crown segmentation polygons derived from multispectral DOP20 imagery across the German federal states of Sachsen (SN, green) and Sachsen-Anhalt (ST, blue), totaling 218.7 million predicted crowns.
(b) Relationship between normalized crown area (x-axis) and estimated crown spread (y-axis) for a subset of the segmented trees from Halle (Saale), ST.}
\end{figure}
```
:::

@@

Validating large-scale tree crown segmentation models reveals a web of interrelated challenges. Among these challenges is phenology, a moving target: the same forest can look drastically different between leaf-on summer imagery and leaf-off winter scenes. Models trained on a single phenological period often struggle in another, yielding inconsistent segmentation accuracy across the year [@sapkota2024; @liu2024; @garnot2025deep]. For example, a canopy delineation that performs well on lush summer foliage may under-segment sparse autumn crowns or miss bare branches in winter (Figure 3a). Such season-driven variability not only degrades model performance but also complicates validation – a one-shot model might appear accurate in one season and fail in the next, raising questions about how and when accuracy should be assessed. Incorporating multi-season data including the key phenological stages: leaf emergence, flowering, fruiting, autumn coloring and leaf fall, during both training and validation is essential, as phenological dynamics have been shown to strongly influence model generalizability [@moussaid2021tree; @cong2022citrus].

@@

::: {.column width="100%"}
```{=latex}
\begin{figure}[htbp]
\centering
\includegraphics[width=0.8\textwidth]{figures/figure2.png}
\caption{Illustration of key validation challenges in large-scale tree crown segmentation across two federal states (Sachsen-Anhalt; in blue \& Sachsen; in green) in Germany. Insets highlight specific issues: (a) seasonal variation (phenology) affecting appearance, (b) tiling artifacts in image preprocessing, (c) misalignment between predicted crowns and known tree stem locations, and (d) discrepancies between predicted crowns (blue) and labeled ground truth (red), emphasizing model performance gaps. Blue (Sachsen-Anhalt) and green (Sachsen) overlays represent modelled individual tree crowns across different Federal States in Germany.}
\end{figure}
```
:::

@@

Validating large-scale tree segmentation models is challenging due to seasonal variability, diverse terrain (Figure 3b,c)[@copernicus_copdem30, @eea_corine2020] and illumination conditions, inconsistent ground truth data, ambiguous evaluation criteria, and scale-dependent performance. These factors significantly affect the reliability and practical value of AI-generated tree maps for forestry, ecology, and urban planning.

@@

\begin{credits}
\subsubsection{\ackname} DeepTrees is part of the \href{https://deeptrees.de}{DeepTrees: Deep-Learning based spatiotemporal tree inventorying and monitoring from public orthoimages} project, funded by the Integration Platform "Sustainable Future Land Use" at Helmholtz-Centre for Environmental Research – UFZ within the Programme oriented Funding (PoF) period IV of the Helmholtz Program "Changing Earth – Sustaining our Future", Topic 5 "Landscapes of the Future. This repository is based on the work described in Freudenberg et al. (2022). This work was supported by Helmholtz Association's Initiative and Networking Fund through Helmholtz AI [grant number: ZT-I-PF-5-01]. This work used resources of the Deutsches Klimarechenzentrum (DKRZ) granted by its Scientific Steering Committee (WLA) under project ID AIM.
